---
title: "Exploring EEG data"
author: "Cameron Roach"
date: "2018-12-29T02:13:00"
slug: eeg-explore
categories: ["R"]
tags: ["R", "EEG"]
description: Exploring EEG data from the Grasp-and-Lift EEG Detection Kaggle competition.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	collapse = TRUE
)

rm(list=ls())

data_dir <- file.path(Sys.getenv("DATA"), "kaggle-eeg")
```

Electroencephalographic (EEG) sensors record electrical activity in the brain. It is a non-invasive technique and has applications in medicine, gaming __TODO__:REF.

## Data

The data is taken from the [Grasp-and-Lift EEG Detection](https://www.kaggle.com/c/grasp-and-lift-eeg-detection) Kaggle competition. 32 channel EEG data was collected from 16 healthy subjects while they conducted several grasp and lift trials.


```{r}
library(tidyverse)

train_df <- data_frame(paths = list.files(file.path(data_dir, "train"),
                                          full.names = TRUE),
                       filename = list.files(file.path(data_dir, "train")),
                       subject = str_extract(filename, "(?<=subj)[[:digit:]]+"),
                       series = str_extract(filename, "(?<=series)[[:digit:]]+"),
                       dtype = str_extract(filename, "(?<=_)[[:alpha:]]+(?=.csv)")) %>% 
  slice(1:10) %>% 
  mutate(data = map(paths, read_csv),
         data = map(data, ~ .x %>% 
                      mutate(id = as.numeric(str_extract(id, "[[:digit:]]+$"))) %>% 
                      rename(ts = id))) %>% 
  select(-c(paths, filename)) %>% 
  spread(dtype, data)

eeg_vars <- names(train_df$data[[1]])[-1]
event_vars <- names(train_df$events[[1]])[-1]
  
train_df <- train_df %>% 
  mutate(data = map2(data, events, inner_join, by = "ts")) %>% 
  select(-events) %>% 
  unnest(data)
```

```{r}
train_df %>% 
  filter(subject == 1,
         series == 1,
         ts <= 1e4) %>% 
  select(ts, !!event_vars) %>% 
  gather(var, val, -ts) %>% 
  mutate(var = factor(var, levels = event_vars, ordered = TRUE)) %>% 
  ggplot(aes(x = ts, y = val)) +
  geom_line() +
  facet_wrap(~var, ncol = 1, strip.position = "right")
```


Wrote my own function to loop through the data and find the start and end indexes of each event. Decided to use `C++` as it was about thirty times faster when I benchmarked against equivalent `R` code and I've been looking for an excuse to play around with `Rcpp`...

```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
DataFrame StartEndIdxsC(NumericVector x, NumericVector ts) {
  int n = x.size();
  NumericMatrix times(n, 2);
  int j = 0;
  int start_idx = 0;
  
  for(int i = 0; i < n; i++) {
    if (start_idx == 0 && x[i] == 1)
    {
      times(j,0) = ts[i];
      start_idx = 1;
    } 
    else if (start_idx == 1 && x[i] == 0) 
    {
      times(j,1) = ts[i];
      start_idx = 0;
      j = j + 1;
    }
  }
  
  NumericMatrix subset(j,2);
  
  for(int i = 0; i < j; i++) {
    subset(i,_) = times(i,_);
  }
  
  DataFrame df = DataFrame::create( Named("start") = subset(_,0),
                                    Named("end") = subset(_,1) );
  
  return df;
}
```



```{r}
# #' Calculates start and end indexes
# #'
# #' NOTE: if final entry is a 1, the end won't be calculated.
# start_end_idxs <- function(x, ts) {
#   n <- length(x)
#   l <- vector("list", n)
#   j <- 1
#   start_idx <- 0
#   
#   for (i in 1:n) {
#     if (start_idx == 0 & x[i] == 1) {
#       start <- ts[i]
#       start_idx <- 1
#     } else if (start_idx == 1 & x[i] == 0) {
#       end <- ts[i]
#       start_idx <- 0
#       l[[j]] <- data_frame(start = start,
#                            end = end)
#       j <- j + 1
#     }
#   }
#   
#   bind_rows(l[!unlist(lapply(l, is.null))])
# }


event_idx <- train_df %>% 
  filter(subject == 1,
         series == 1,
         HandStart == 1) %>% 
  pull(ts) %>% 
  min()

start_buffer <- 500  # 500 Hz sampling
end_buffer <- 3000

plot_eeg_df <- train_df %>% 
  filter(subject == 1,
         series == 1,
         between(ts, event_idx - start_buffer, event_idx + end_buffer)) %>% 
  select(ts, !!eeg_vars) %>% 
  gather(sensor, val, -ts) %>% 
  mutate(type = str_extract(sensor, "[[:alpha:]]+"))

plot_event_df <- train_df %>% 
  filter(subject == 1,
         series == 1,
         between(ts, event_idx - start_buffer, event_idx + end_buffer)) %>% 
  select(ts, !!event_vars) %>% 
  gather(event, val, -ts) %>% 
  group_by(event) %>%
  nest() %>% 
  mutate(times = map(data, ~ StartEndIdxsC(.$val, .$ts))) %>% 
  select(event, times) %>% 
  mutate(event = factor(event, levels = event_vars, ordered = TRUE)) %>% 
  unnest()

ggplot() +
  geom_line(data = plot_eeg_df,
            aes(x = ts, y = val, group = sensor)) +
  geom_rect(data = plot_event_df,
            aes(xmin=start, xmax=end, ymin=-Inf, ymax=Inf, fill = event),
            alpha = 0.5) +
  facet_wrap(~type) +
  theme(legend.position = "none")
```


# TODO

* Combine event data so that it is just two columns. ts and event. Check that events don't occur at the same time.
* Once this is done, plot all events as vertical lines in the above plot.